---
title: "HVD Exploratory Analysis for County Disaggregation <img src=\"HPIO_Logo_Small.jpg\" style=\"float: right;\"/>" 
author: "Lareina La Flair, MPH, PhD, Health Policy Institute of Ohio"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_collapsed: true
    number_sections: true
    theme: lumen
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

packages <- c("dplyr", "forcats", "plotly", "stringr", "tidyr", "data.table", "purrr", "kableExtra", "flextable", "plotly", "unikn", "leaflet", "leaflegend", "leaflet.extras", "sf", "censusapi", "tidycensus", "tigris", "htmlwidgets")


# Install packages not yet installed
installed_packages <- packages %in% rownames(installed.packages())
if (any(installed_packages == FALSE)) {
  install.packages(packages[!installed_packages])
}
# Packages loading
invisible(lapply(packages, library, character.only = TRUE))

```

# Background

HPIO has an emerging need to produce *choropleth maps* for visualizing an aggregate summary of some condition or outcome (e.g., population density, percent Black race, social vulnerability, suicide deaths) across Ohio counties for the Health Value Dashboard. 

For the 2023 Health Value Dashboard, we are tasked with identifying metrics for county disaggregation. The metrics below are Ohio-specific and are intended to be presented jointly, pairing one community condition measure with a health outcome at the county level. **This analysis is exploratory.**


# Make the maps

Read in the dataset with the health event or condition you're looking to map.

Here we are using  Social Vulnerability Index by county from the Ohio Department of Health [Public Health Data Warehouse](https://odh.ohio.gov/know-our-programs/informatics-and-data-management/resources/odh-full-banner-public-data-warehouse)

(After cleaning) The data have the following columns: 
* `county` as Ohio county of residence
* `index` expressed as number of deaths per 100K population


```{r read-health-data, echo=FALSE, eval=TRUE, warning = FALSE, include = FALSE}
## Map table is a dataframe with the count of OD (or the health event of interest) aggregated by county

svi <- fread("./data/Ohio_COUNTY_SVI_2020.csv")
names(svi)

#clean up columns in your dataset 

map_table_svi  <- svi %>% 
  #filter(DeathFentanylAnaloguesIndFentanylAnalogues == "Yes" &  DeathCountyCountyName != "Total") %>% 
  rename(county = COUNTY, svi = RPL_THEMES) %>% 
  select(FIPS, county, E_TOTPOP, svi)

map_table_svi$svi <- round(map_table_svi$svi, digits = 3)
map_table_svi$county <- tolower(map_table_svi$county)

map_table_svi %>% 
  arrange(desc(svi))

## Infant Mortality Rate


imr <- fread("./data/OH_county_IMR_2017-2021.csv")
names(imr)

#clean up columns in your dataset 

map_table_imr <- imr %>% 
  filter(DeathCountyCountyName != "Total") %>% 
  rename(county = DeathCountyCountyName, count = Deaths, rate = DeathRate) %>% 
  select(county, rate, count)

map_table_imr$rate <- round(map_table_imr$rate, digits = 3)
map_table_imr$county <- tolower(map_table_imr$county)

map_table_imr %>% 
  arrange(desc(count))

## Low Birth Weight

lbw <- fread("./data/lbw_2014-2020.csv")
names(lbw)

#clean up columns in your dataset 

map_table_lbw <- lbw %>% 
  rename(percent = value) %>% 
  select(county, percent)

map_table_lbw$percent <- round(map_table_lbw$percent, digits = 3)
map_table_lbw$county <- tolower(map_table_lbw$county)

map_table_lbw %>% 
  arrange(desc(percent))

## Food insecurity


fins <- fread("./data/fins_2019.csv")
names(fins)

#clean up columns in your dataset 

map_table_fins <- fins %>% 
  rename(percent = value) %>% 
  select(county, percent)

map_table_fins$percent <- round(map_table_fins$percent, digits = 3)
map_table_fins$county <- tolower(map_table_fins$county)

map_table_fins%>% 
  arrange(desc(percent))

## Mental health providers


mhprov <- fread("./data/mhprovider_2021.csv")
names(mhprov)

#clean up columns in your dataset 

map_table_mhprov<- mhprov %>% 
  rename(county = County, count = provider_count) %>% 
  select(county, ratio, count)

# remove commas and switch to lowercase
map_table_mhprov$count <- as.numeric(gsub(",","",map_table_mhprov$count))
map_table_mhprov$ratio <- (gsub(",","",map_table_mhprov$ratio))
map_table_mhprov$county <- tolower(map_table_mhprov$county)


map_table_mhprov$ratio <- str_extract(map_table_mhprov$ratio, "[0-9]+")


map_table_mhprov%>% 
  arrange(desc(ratio))

## Suicide deaths

suic <- fread("./data/suic_county_2020-2021.csv")
names(suic)

#clean up columns in your dataset 

map_table_suic <- suic %>% 
  rename(county = DeathCountyCountyName, count = Deaths, rate = DeathRate) %>% 
  select(county, count, rate)

# switch to lowercase
map_table_suic$county <- tolower(map_table_suic$county)


map_table_suic%>% 
  arrange(desc(rate))

## Overdose deaths

od <- fread("./data/od_county_2020-2021.csv")
names(od)

#clean up columns in your dataset 

map_table_od <- od %>% 
  rename(county = DeathCountyCountyName, count = Deaths, rate = DeathRate) %>% 
  select(county, count, rate)

# switch to lowercase
map_table_od$county <- tolower(map_table_od$county)


map_table_od%>% 
  arrange(desc(rate))


```

## Load the shape files

Here, we read in the shape file for counties, filtered to Ohio counties. Centroids are the mathematical center of a geographic unit (here, counties).

```{r read-shape-file, echo=FALSE, eval=TRUE, warning = FALSE, include = FALSE}
## Load shape file from Tigris
counties <- counties("OH", cb=T)
centroids <- counties %>%
  select(NAME, geometry) %>% # select only the necessary columns
  sf::st_centroid(NAME)  

## check out the unadorned map quickly (eww, those blue lines!)

counties %>% 
  leaflet() %>% 
  addTiles() %>% 
  addPolygons(popup=~NAME)

```

## Spatial joins

Spatially join the outcome (here, deaths) to the geographic unit (here, counties).


```{r spatial-join, echo=FALSE, eval=TRUE, warning = FALSE, include = FALSE}
## spatial join visits to counties

counties$NAME <- tolower(counties$NAME)

#svi

county_merged_svi <- geo_join(counties, map_table_svi, "NAME", "county")
county_merged_svi <- subset(county_merged_svi, !is.na(count))

#imr
county_merged_imr <- geo_join(counties, map_table_imr, "NAME", "county")
county_merged_imr <- subset(county_merged_imr, !is.na(count))

#lbw
county_merged_lbw <- geo_join(counties, map_table_lbw, "NAME", "county")
county_merged_lbw <- subset(county_merged_lbw, !is.na(percent))

#fins
county_merged_fins <- geo_join(counties, map_table_fins, "NAME", "county")
county_merged_fins <- subset(county_merged_fins, !is.na(percent))

#mhprovider

county_merged_mhprov <- geo_join(counties, map_table_mhprov, "NAME", "county")
county_merged_mhprov <- subset(county_merged_mhprov, !is.na(ratio))

#suicide

county_merged_suic <- geo_join(counties, map_table_suic , "NAME", "county")
county_merged_suic  <- subset(county_merged_suic, !is.na(rate))

#overdose
#suicide

county_merged_od <- geo_join(counties, map_table_od , "NAME", "county")
county_merged_od  <- subset(county_merged_od, !is.na(rate))



```

## Basic interactive maps

``
Set up color palettes, popups, and base layers -- putting it all together using the [`leaflet` package](https://rstudio.github.io/leaflet/).


```{r display-unadjusted, echo=FALSE, eval=TRUE, warning = FALSE, include = FALSE}

### Basic maps ###

## svi

# Creating a color palette based on the number range 

mypalette <- colorNumeric(palette="YlGnBu", domain=county_merged_svi$svi, na.color="transparent")
pal_quan <- colorQuantile(palette = "YlGnBu", n = 4, domain = county_merged_svi$svi)

popup_county_svi <- paste0("<strong>", county_merged_svi$NAME, 
                       "</strong><br />Social Vulnerability Index: ",
                       as.character(county_merged_svi$svi))


svi_map <- county_merged_svi %>%
  leaflet() %>%
  addProviderTiles("CartoDB.Positron") %>%
  setView(-82.983330, 39.983334, zoom = 7) %>% #set long and lat coordinates to Columbus as state center
  addPolygons(data = county_merged_svi, 
             fillColor = ~pal_quan(county_merged_svi$svi), 
             fillOpacity = 0.7, 
             weight = 0.2, 
             smoothFactor = 0.2, 
             popup = ~popup_county_svi) %>%
  # Add some labels to counties
  addLabelOnlyMarkers(
    data = centroids,
    label = ~NAME,
    labelOptions = labelOptions(noHide = TRUE, direction = 'center', textOnly = TRUE, opacity = 0.3)) %>% 
   addLegend(pal = pal_quan, 
          values = county_merged_svi$svi, 
          position = "bottomleft", 
          title = "Social Vulnerability Index, 2020, Quantiles") %>% 
  leaflet.extras::setMapWidgetStyle(list(background= "transparent")) %>%
   leaflet.extras::addResetMapButton()

svi_map

## imr

# Creating a color palette based on the number range 

mypalette <- colorNumeric(palette="YlGnBu", domain=county_merged_imr$rate, na.color="transparent")
pal_quan <- colorQuantile(palette = "YlGnBu", n = 4, domain = county_merged_imr$rate)

popup_county_imr <- paste0("<strong>", county_merged_imr$NAME, 
                       "</strong><br />Infant Mortality Rate (deaths per 100K pop): ",
                       as.character(county_merged_imr$rate))


imr_map <- county_merged_imr %>%
  leaflet() %>%
  addProviderTiles("CartoDB.Positron") %>%
  setView(-82.983330, 39.983334, zoom = 7) %>% #set long and lat coordinates to Columbus as state center
  addPolygons(data = county_merged_imr, 
             fillColor = ~pal_quan(county_merged_imr$rate), 
             fillOpacity = 0.7, 
             weight = 0.2, 
             smoothFactor = 0.2, 
             popup = ~popup_county_imr) %>%
  # Add some labels to counties
  addLabelOnlyMarkers(
    data = centroids,
    label = ~NAME,
    labelOptions = labelOptions(noHide = TRUE, direction = 'center', textOnly = TRUE, opacity = 0.3)) %>% 
   addLegend(pal = pal_quan, 
          values = county_merged_imr$rate, 
          position = "bottomleft", 
          title = "Infant Mortality Rate, 2017-2021, Quantiles") %>% 
  leaflet.extras::setMapWidgetStyle(list(background= "transparent")) %>%
   leaflet.extras::addResetMapButton()

imr_map

## lbw

# Creating a color palette based on the number range 

mypalette <- colorNumeric(palette="YlGnBu", domain=county_merged_lbw$percent, na.color="transparent")
#pal_quan <- colorQuantile(palette = "YlGnBu", n = 4, domain = county_merged_lbw$percent)

popup_county_lbw <- paste0("<strong>", county_merged_lbw$NAME, 
                       "</strong><br />Low birthweight (Percent of births <2500 g): ",
                       as.character(county_merged_lbw$percent))


lbw_map <- county_merged_lbw %>%
  leaflet() %>%
  addProviderTiles("CartoDB.Positron") %>%
  setView(-82.983330, 39.983334, zoom = 7) %>% #set long and lat coordinates to Columbus as state center
  addPolygons(data = county_merged_lbw, 
             fillColor = ~mypalette(county_merged_lbw$percent), 
             fillOpacity = 0.7, 
             weight = 0.2, 
             smoothFactor = 0.2, 
             popup = ~popup_county_lbw) %>%
  # Add some labels to counties
  addLabelOnlyMarkers(
    data = centroids,
    label = ~NAME,
    labelOptions = labelOptions(noHide = TRUE, direction = 'center', textOnly = TRUE, opacity = 0.3)) %>% 
   addLegend(pal = mypalette, 
          values = county_merged_lbw$percent, 
          position = "bottomleft", 
          title = "Percent Low Birth Weight, 2016-2020") %>% 
  leaflet.extras::setMapWidgetStyle(list(background= "transparent")) %>%
   leaflet.extras::addResetMapButton()

lbw_map

## fins

# Creating a color palette based on the number range 

mypalette <- colorNumeric(palette="YlGnBu", domain=county_merged_fins$percent, na.color="transparent")
pal_quan <- colorQuantile(palette = "YlGnBu", n = 4, domain = county_merged_fins$percent)

popup_county_fins <- paste0("<strong>", county_merged_fins$NAME, 
                       "</strong><br />Percent of Population Lacking Adequate Access to Food: ",
                       as.character(county_merged_fins$percent))


fins_map <- county_merged_fins %>%
  leaflet() %>%
  addProviderTiles("CartoDB.Positron") %>%
  setView(-82.983330, 39.983334, zoom = 7) %>% #set long and lat coordinates to Columbus as state center
  addPolygons(data = county_merged_fins, 
             fillColor = ~pal_quan(county_merged_fins$percent), 
             fillOpacity = 0.7, 
             weight = 0.2, 
             smoothFactor = 0.2, 
             popup = ~popup_county_fins) %>%
  # Add some labels to counties
  addLabelOnlyMarkers(
    data = centroids,
    label = ~NAME,
    labelOptions = labelOptions(noHide = TRUE, direction = 'center', textOnly = TRUE, opacity = 0.3)) %>% 
   addLegend(pal = pal_quan, 
          values = county_merged_fins$percent, 
          position = "bottomleft", 
          title = "Percent Food Insecure, 2019, Quantiles") %>% 
  leaflet.extras::setMapWidgetStyle(list(background= "transparent")) %>%
   leaflet.extras::addResetMapButton()

fins_map

## mh provider


popup_county_mhprovider <- paste0("<strong>", county_merged_mhprov$NAME,"</strong><br />Ratio of population to mental health providers: ",as.character(county_merged_mhprov$ratio))

county_merged_mhprov$ratio <- as.numeric(county_merged_mhprov$ratio)

pal <- colorBin(
  palette = "Reds"
  , domain = county_merged_mhprov$ratio
  , bins = 8
 
)

pal_quan <- colorQuantile(palette = "Reds", n = 4, domain = county_merged_mhprov$ratio)

mhprov_map <- county_merged_mhprov %>%
  leaflet() %>%
  addProviderTiles("CartoDB.Positron") %>%
  setView(-82.983330, 39.983334, zoom = 7) %>% #set long and lat coordinates to Columbus as state center
  addPolygons(data = county_merged_mhprov, 
             fillColor = ~pal_quan(county_merged_mhprov$ratio), 
             fillOpacity = 0.7, 
             weight = 0.2, 
             smoothFactor = 0.2, 
             popup = ~popup_county_mhprovider) %>%
  # Add some labels to counties
  addLabelOnlyMarkers(
    data = centroids,
    label = ~NAME,
    labelOptions = labelOptions(noHide = TRUE, direction = 'center', textOnly = TRUE, opacity = 0.3)) %>% 
   addLegend(pal = pal_quan, 
          values = county_merged_mhprov$ratio, 
          position = "bottomleft", 
          title = "Ratio: population to mental health providers<br />per 100,000 population<br/> 2021, Quantiles") %>% 
  leaflet.extras::setMapWidgetStyle(list(background= "transparent")) %>%
   leaflet.extras::addResetMapButton()

mhprov_map

## suicide deaths

popup_county_suic <- paste0("<strong>", county_merged_suic$NAME,"</strong><br />Suicide Death Rate: ",as.character(county_merged_suic$rate))

county_merged_suic$rate <- as.numeric(county_merged_suic$rate)

pal <- colorBin(
  palette = "Reds"
  , domain = county_merged_suic$rate
  , bins = 8
 
)

pal_quan <- colorQuantile(palette = "Reds", n = 4, domain = county_merged_suic$rate)

suic_map <- county_merged_suic %>%
  leaflet() %>%
  addProviderTiles("CartoDB.Positron") %>%
  setView(-82.983330, 39.983334, zoom = 7) %>% #set long and lat coordinates to Columbus as state center
  addPolygons(data = county_merged_suic, 
             fillColor = ~pal_quan(county_merged_suic$rate), 
             fillOpacity = 0.7, 
             weight = 0.2, 
             smoothFactor = 0.2, 
             popup = ~popup_county_suic) %>%
  # Add some labels to counties
  addLabelOnlyMarkers(
    data = centroids,
    label = ~NAME,
    labelOptions = labelOptions(noHide = TRUE, direction = 'center', textOnly = TRUE, opacity = 0.3)) %>% 
   addLegend(pal = pal_quan, 
          values = county_merged_suic$rate, 
          position = "bottomleft", 
          title = "Number of Suicides<br />per 100,000 population<br/>2020-2021, Quantiles") %>% 
  leaflet.extras::setMapWidgetStyle(list(background= "transparent")) %>%
   leaflet.extras::addResetMapButton()

suic_map

## overdose deaths

popup_county_od <- paste0("<strong>", county_merged_suic$NAME,"</strong><br />Overdose Death Rate: ",as.character(county_merged_od$rate))

county_merged_od$rate <- as.numeric(county_merged_od$rate)

pal <- colorBin(
  palette = "Reds"
  , domain = county_merged_od$rate
  , bins = 8
 
)

pal_quan <- colorQuantile(palette = "Reds", n = 4, domain = county_merged_od$rate)

od_map <- county_merged_od %>%
  leaflet() %>%
  addProviderTiles("CartoDB.Positron") %>%
  setView(-82.983330, 39.983334, zoom = 7) %>% #set long and lat coordinates to Columbus as state center
  addPolygons(data = county_merged_od, 
             fillColor = ~pal_quan(county_merged_od$rate), 
             fillOpacity = 0.7, 
             weight = 0.2, 
             smoothFactor = 0.2, 
             popup = ~popup_county_od) %>%
  # Add some labels to counties
  addLabelOnlyMarkers(
    data = centroids,
    label = ~NAME,
    labelOptions = labelOptions(noHide = TRUE, direction = 'center', textOnly = TRUE, opacity = 0.3)) %>% 
   addLegend(pal = pal_quan, 
          values = county_merged_od$rate, 
          position = "bottomleft", 
          title = "Number of Unintentional Overdose Deaths<br />per 100,000 population<br/> 2020-2021, Quantiles") %>% 
  leaflet.extras::setMapWidgetStyle(list(background= "transparent")) %>%
   leaflet.extras::addResetMapButton()

od_map

#save maps as html files

saveWidget(mhprov_map, file = "mhprov_map.html")
saveWidget(od_map, file = "od_map.html")
saveWidget(suic_map, file = "suic_map.html")
saveWidget(lbw_map, file = "lbw_map.html")
saveWidget(imr_map, file = "imr_map.html")
saveWidget(fins_map, file = "fins_map.html")

```


# Social Vulnerability Index Maps {.tabset .tabset-pills}

Pairings with Social Vulnerability Index:

* SVI and Infant Mortality by County
* SVI and Low Birth Weight by County
* SVI and Food Insecurity by County

## Pairing 1: SVI and Infant Mortality by County

**Data sources:**

**Social Vulnerability Index**

Social Vulnerability Index by county comes from the Ohio Department of Health [Public Health Data Warehouse](https://odh.ohio.gov/know-our-programs/informatics-and-data-management/resources/odh-full-banner-public-data-warehouse)
[**Social Vulnerability Index (SVI)**](https://odh.ohio.gov/know-our-programs/health-equity/health-improvement-zones)created by the US Centers for Disease Control and Prevention (CDC), using the most current data available from the US Census Bureau American Community Survey 5-year estimates (2014-2018). The SVI is a **score ranging from 0 – 1** and is comprised of 15 indicators grouped into 4 themes: Socioeconomic Status (Below Poverty, Unemployed, Income, No High School Diploma), Household Composition & Disability (Aged 65 and Older, Aged 17 and Younger, Civilian with a Disability, Single-Parent Households), Minority Status & Language (Minority, Speak English “Less than Well”), Housing & Transportation (Multi-Unit Structures, Mobile Homes, Crowding, No Vehicle, Group Quarters)

```{r svi_imr, echo=FALSE, eval=TRUE, warning = FALSE, include = TRUE}

svi_map
imr_map

```

## Pairing 2: SVI and Low Birth Weight by County

**Data sources:**

**County Health Rankings** 
Percentage of live births with low birthweight (< 2,500 grams). The 2022 County Health Rankings used data from 2014-2020 for this measure.

[Low birth weight](https://www.countyhealthrankings.org/explore-health-rankings/county-health-rankings-model/health-outcomes/quality-of-life/low-birthweight?year=2022&state=39&tab=1)


**Social Vulnerability Index**

Social Vulnerability Index by county comes from the Ohio Department of Health [Public Health Data Warehouse](https://odh.ohio.gov/know-our-programs/informatics-and-data-management/resources/odh-full-banner-public-data-warehouse)
**Social Vulnerability Index (SVI)** created by the US Centers for Disease Control and Prevention (CDC), using the most current data available from the US Census Bureau American Community Survey 5-year estimates (2014-2018). The SVI is a **score ranging from 0 – 1** and is comprised of 15 indicators grouped into 4 themes: Socioeconomic Status (Below Poverty, Unemployed, Income, No High School Diploma), Household Composition & Disability (Aged 65 and Older, Aged 17 and Younger, Civilian with a Disability, Single-Parent Households), Minority Status & Language (Minority, Speak English “Less than Well”), Housing & Transportation (Multi-Unit Structures, Mobile Homes, Crowding, No Vehicle, Group Quarters)

```{r svi_lbw, echo=FALSE, eval=TRUE, warning = FALSE, include = TRUE}

svi_map

lbw_map

```

## Pairing 3: SVI and Food Insecurity

**Data source:**

**County Health Rankings** 

Percentage of population who lack adequate access to food. The 2022 County Health Rankings used data from 2019 for this measure.

[Food insecurity](https://www.countyhealthrankings.org/explore-health-rankings/county-health-rankings-model/health-factors/health-behaviors/diet-and-exercise/food-insecurity?year=2022&state=39&tab=1)


**Social Vulnerability Index**

Social Vulnerability Index by county comes from the Ohio Department of Health [Public Health Data Warehouse](https://odh.ohio.gov/know-our-programs/informatics-and-data-management/resources/odh-full-banner-public-data-warehouse)
**Social Vulnerability Index (SVI)** created by the US Centers for Disease Control and Prevention (CDC), using the most current data available from the US Census Bureau American Community Survey 5-year estimates (2014-2018). The SVI is a **score ranging from 0 – 1** and is comprised of 15 indicators grouped into 4 themes: Socioeconomic Status (Below Poverty, Unemployed, Income, No High School Diploma), Household Composition & Disability (Aged 65 and Older, Aged 17 and Younger, Civilian with a Disability, Single-Parent Households), Minority Status & Language (Minority, Speak English “Less than Well”), Housing & Transportation (Multi-Unit Structures, Mobile Homes, Crowding, No Vehicle, Group Quarters)

```{r svi_fins, echo=FALSE, eval=TRUE, warning = FALSE, include = TRUE}

svi_map

fins_map

```


# Mental Health Provider Maps {.tabset .tabset-pills}

Pairings with Mental Health Providers:

* Mental Health Providers and Suicide Deaths by County
* Mental Health Providers and Overdose Deaths by County

## Pairing 4: Mental Health Providers and Suicide Deaths by County

**Data sources:**

**County Health Rankings:**

[County Health Ranking Mental Health Providers](https://www.countyhealthrankings.org/explore-health-rankings/county-health-rankings-model/health-factors/clinical-care/access-to-care/mental-health-providers?year=2022): Ratio of the population to mental health providers. The ratio represents the number of individuals served by one mental health provider in a county. Ratio of population to mental health providers. The 2022 County Health Rankings used data from *2021* for this measure.

Access to care requires not only financial coverage, but also access to providers. Nearly thirty percent of the population lives in a county designated as a Mental Health Professional Shortage Area. As the mental health parity aspects of the Affordable Care Act create increased coverage for mental health services, many anticipate increased workforce shortages. 

Mental Health Providers is the **ratio of the population to mental health providers**. The ratio represents the number of individuals served by one mental health provider in a county, if the population was equally distributed across providers. For example, if a county has a population of 50,000 and has 20 mental health providers, their ratio would be: 2,500:1. 

**Ohio Department of Health Public Health Warehouse**

*Filters applied:*

by County
for County = (Adams, Allen, Ashland, Ashtabula, Athens, Auglaize, Belmont, Brown, Butler, Carroll, Champaign, Clark, Clermont, Clinton, Columbiana, Coshocton, Crawford, Cuyahoga, Darke, Defiance, Delaware, Erie, Fairfield, Fayette, Franklin, Fulton, Gallia, Geauga, Greene, Guernsey, Hamilton, Hancock, Hardin, Harrison, Henry, Highland, Hocking, Holmes, Huron, Jackson, Jefferson, Knox, Lake, Lawrence, Licking, Logan, Lorain, Lucas, Madison, Mahoning, Marion, Medina, Meigs, Mercer, Miami, Monroe, Montgomery, Morgan, Morrow, Muskingum, Noble, Ottawa, Paulding, Perry, Pickaway, Pike, Portage, Preble, Putnam, Richland, Ross, Sandusky, Scioto, Seneca, Shelby, Stark, Summit, Trumbull, Tuscarawas, Union, Van Wert, Vinton, Warren, Washington, Wayne, Williams, Wood, Wyandot) , External Injury Intent = (Suicide) , Year = (2020, 2021 **) , Res State = (OH) 

2021 and 2020 pooled due to low numbers.

```{r mhprov_suic, echo=FALSE, eval=TRUE, warning = FALSE, include = TRUE}

mhprov_map
suic_map

```

## Pairing 5: Mental Health Providers and Overdose Deaths by County

**Data Source**

**County Health Rankings:**

[County Health Ranking Mental Health Providers](https://www.countyhealthrankings.org/explore-health-rankings/county-health-rankings-model/health-factors/clinical-care/access-to-care/mental-health-providers?year=2022): Ratio of the population to mental health providers. The ratio represents the number of individuals served by one mental health provider in a county. Ratio of population to mental health providers. The 2022 County Health Rankings used data from *2021* for this measure.

Access to care requires not only financial coverage, but also access to providers. Nearly thirty percent of the population lives in a county designated as a Mental Health Professional Shortage Area. As the mental health parity aspects of the Affordable Care Act create increased coverage for mental health services, many anticipate increased workforce shortages. 

Mental Health Providers is the **ratio of the population to mental health providers**. The ratio represents the number of individuals served by one mental health provider in a county, if the population was equally distributed across providers. For example, if a county has a population of 50,000 and has 20 mental health providers, their ratio would be: 2,500:1. 

**Ohio Department of Health Public Health Warehouse**

*Filters applied:*

by County
for County = (Adams, Allen, Ashland, Ashtabula, Athens, Auglaize, Belmont, Brown, Butler, Carroll, Champaign, Clark, Clermont, Clinton, Columbiana, Coshocton, Crawford, Cuyahoga, Darke, Defiance, Delaware, Erie, Fairfield, Fayette, Franklin, Fulton, Gallia, Geauga, Greene, Guernsey, Hamilton, Hancock, Hardin, Harrison, Henry, Highland, Hocking, Holmes, Huron, Jackson, Jefferson, Knox, Lake, Lawrence, Licking, Logan, Lorain, Lucas, Madison, Mahoning, Marion, Medina, Meigs, Mercer, Miami, Monroe, Montgomery, Morgan, Morrow, Muskingum, Noble, Ottawa, Paulding, Perry, Pickaway, Pike, Portage, Preble, Putnam, Richland, Ross, Sandusky, Scioto, Seneca, Shelby, Stark, Summit, Trumbull, Tuscarawas, Union, Van Wert, Vinton, Warren, Washington, Wayne, Williams, Wood, Wyandot) , External Injury Intent = (Unintentional) , External Injury Mechanism Group = (Poisoning) , Year = (2020, 2021 **) , Res State = (OH) 

2021 and 2020 pooled due to low numbers.


```{r mhprov_od, echo=FALSE, eval=TRUE, warning = FALSE, include = TRUE}

mhprov_map
od_map
```

# Correlation analysis

Positive correlations are displayed in blue and negative correlations in red color. Color intensity and the size of the circle are proportional to the correlation coefficients. In the right side of the correlogram, the legend color shows the correlation coefficients and the corresponding colors.

**Interpretation**
Mental health provider ratio appears to correlate moderately with county suicide rate *but* not in the expected direction. Under the assumption that a higher concentration of providers is associated with greater access to care and, ultimately, reduced suicide risk, we would expect a higher population-to-provider ratio to be positively correlated with suicide rates (higher the ratio, the higher the rate), but here we observe a negative correlation (higher the ratio, the lower the rate). This suggests that other factors operating at the county-, community-, and household- and individual-level may be at play.

```{r corr, echo=FALSE, eval=TRUE, warning=FALSE, message=FALSE, results='hide'}
#join df together by county

#put all data frames into list
df_list <- list(map_table_mhprov, map_table_od, map_table_suic)

#merge all data frames in list
mh <- df_list %>% 
  reduce(left_join, by='county') 


mh <- mh %>% 
  rename(suicrt = rate.x, odrt = rate.y, suict = count.x, odct = count.y) %>% 
  select(suicrt, odrt, ratio, suict, odct)

mh$suicrt <- as.numeric(mh$suicrt)
mh$odrt <- as.numeric(mh$odrt)  
mh$ratio <- as.numeric(mh$ratio) 

library(corrplot)

m <- cor(mh)
#m <- cor.mtest(mh, conf.level = 0.95)
invisible(m)

corr_circle <- corrplot(m, method = "circle" , type = "upper", order = "hclust", 
         tl.col = "black", tl.srt = 45)
corr_circle

corr_numeric <- corrplot(m, method = "number" , type = "upper", order = "hclust", 
         tl.col = "black", tl.srt = 45)
corr_numeric


```


End transmission!
